---
author: Yves Deville
title: "Introduction to the `dailymet`package"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
linkcolor: blue
---

# Goals

- Use daily meteorological timeseries, mainly `TX`. 

- Provide bases of functions of the date: trigonometric, polynomial, ...

- Fit time-varying and non-stationary *POT models* with emphasis on
the sensitivity to the *threshold choice*, thanks to dedicated classes
of objects that we may call "TLists" for *list by threshold*.


# The `dailyMet`class

The package comes with the `Rennes` example.

```{r Rennes, message=FALSE} 
library(dailymet)
class(Rennes)
Rennes
methods(class = "dailyMet")
```

So `Rennes` has class `"dailyMet"` which inherits from the
`"data.frame"`class. The `print` method provides useful information. A
method which is not implemented for the class `"dailyMet"` will be
inherited from the `"data.frame"` class e.g., `head`, `tail`, `nrow`
...


```{r Rennes2, message=FALSE} 
autoplot(Rennes)

```

Since the object embeds several variables related to the date, we can
use these straightforwardly. Mind that the variable names are
capitalised.

```{r Rennes3, message=FALSE} 
autoplot(Rennes, subset = Year >= 2010 & JJA, group = "year")

```
Splitting the timeseries in years is suitable when the interest is on
the summer season. It the interest is instead on winter, then using 
the "winter year" `YearW` will be better.


```{r Rennes4, message=FALSE} 
autoplot(Rennes, subset = Year >= 2010, group = "yearW")

```

Although no other data is shipped with the package, it has a sketch 
of file-based database based on the information provided in 
`stationsMF`

```{r stations, message=FALSE} 
findStationMF("troy")
```

The idea is that the name or description can be quite loose. The
reliable identifier of a station is given in `Code`, which corresponds
to Météo-France [données publiques
Météo-France](https://donneespubliques.meteofrance.fr/)


```{r stations2, message=FALSE} 
head(stationsMF)
```

```{r stations3, echo=FALSE, message=FALSE} 
library(leaflet)
m <- leaflet() %>% addTiles;
## m <- m %>% fitBounds(lng1 = -4.5, lat1 = 41.9, lng2 = 7.9, lat2 = 50.8)
m <- m %>% setView(2, 47.1, zoom = 5)
m <- m %>% addMarkers(lng = stationsMF[ , "Lon"],
lat = stationsMF[ , "Lat"],
popup = stationsMF[ , "ShortName"])
m
```

See the help `stationsMF` for a **leaflet** map of these. Note that
the codes of the main MF stations for the metropolitan area correspond
rougthly to an ordering from North to South (decreasing latitude) and
West to East (increasing longitude). This can be of some help.

# Using *TList*s

### What are TLists?

A T list is a list of object sharing the same definition except for
their threshold. The threshold is related to a probability `tau`
corresponding to the probability  $\tau$ as used in quantile regression.

- A `rqTList` object is a list of `rq` objects differing only by their
probability `tau`, hence related to the same data, same formula, ...

- A `fevdTList`object is a list of `fevd` objects representing
  Poisson-GP POT models differing *only by their threshold* which is
  given by quantile regression and corresponds to different
  probabilities `tau`. Again the `fevd`objects are related to the same
  data, to the same formulas, ...
  
Implicitly assumed, the vector `tau` should be in strictly increasing
order.

### Some important remarks

- A `rq` object does not encode a full distribution nor even a tail
  distribution. The `rq` objects may fail to be consistent and the
  quantile may fail to increase with the proability $\tau$. See XXX
  Northrop.

- Stricly speaking, the `fevd` objects in a `fevdTList` object are
  *different* Poisson-GP models, although they must have the same
  formulas. Indeed, the Poisson-GP parameterisation is not independent
  of the threshold: if the GPD scale depends linearly of a covariate
  $x$ for a given threshold $u$, then for a higher threshold $u'>u$
  the scale will depend on $x$ in a nonlinear fashion. See the
  relation between the two POT parameterisation Poisson-GP and NHPP,
  only the later being independent of the threshold.

### Example
